Seismic parameter estimation and the Canadian crust

### 2013-09-11 paper/draftBP.tex
Composition plots

Mean crustal composition is an important parameter in the literature studying the continental crust. Mean composition should not be confused as an estimate of real physical crustal lithological and mineral assemblages. Rather, mean crustal composition is a representative quantity that is useful for constraining crustal genesis and evolution in space and time. Early attempts at estimating mean crustal composition were largely accomplished through extrapolations of upper crustal rocks [Rudnick and Gao, 2003]. After the plate-tectonic revolution, Taylor [1967, 1977], developed the 'island arc' or 'andesite' model which posited that bulk crustal composition should be equivalent to the average convergent-margin andesite. Several problems with this theory, such as the fact that island arcs are more basaltic than andesitic and therefore produce basaltic crustal accretions and cannot account for the intermediate composition of post-Archean crust, led to the introduction of seismic methods to estimate mid and lower crustal composition [Christensen and Mooney, 1995; Rudnick and Fountain, 1995]. This study differs from these later efforts in that it follows more closely to the complete crustal seismic analysis in the work of Smithson et. al. [1981].

@@

The mean seismic velocity ratio computed in this study is combined with values for $V_P$ from the Crust 1.0 model to enable estimates for $V_P$ and $V_S$ to be used to estimate mean crustal composition. Figure \ref{fig:lith} shows mineral assemblages and weighted Canadian Shield estimates plotted against $V_P$ and $V_S$. Mineral assemblages are chosen as useful representations of common crustal rocks and, specifically, for crust in Shield-like regions. Examples of common shield lithologies can be seen in the 30km of paleo-depth represented in the Kapuskasing uplift in the Superior province [Percival and West, 1994]. This uplift is largely comprised of Archean granite-greenstone terranes dominated by granitoid and gneissic rocks, and contains supracrustal assemblages generally metamorphosed to greenschist to amphibolite facies (hence the inclusion of hornblendite and amphibolite) and early plutonic structures comprised largely of tonalites. Tonalites and granodiorites are also part of the TGG suite (for tonalite-trondhjemite-granodiorite) which dominate Archaean cratons [Rudnick, 1995]. Seismic velocity values for these lithologies are taken from the work of Christensen [1996] and are selected at average crustal pressures of 600 MPa. Lines of constant Poisson's Ratio are plotted for reference.

The Slave, Superior and Churchill provinces share similar mean crustal compositions, with the Slave province being slightly more tonalitic in average composition. These three provinces cluster around the average for the Canadian Shield. The Grenville Province on the other hand differs significantly in mean composition. One interpretation is that the crust in the Grenville province has a mean crustal composition corresponding to a more mafic lower crustal composition than other Shield provinces. This interpretation is consistent with a tectonic model for the the central Grenville Province which suggests that the mid-crust has been completely excised locally [Eaton et. al., 2005; Rivers, 2008]. However, many stations in the South-East of the province also have seismic ratios greater than 1.78, similar to the high values common in the adjacent St. Lawrence Platform (where the geometric mean seismic ratio is 1.79).

 [Any other reasons why Grenville and St. Lawrence Platform have higher R values than rest of Shield???]

[ Continue with crustal composition discussion ]
[ Note: I should include Platforms and Orogens in this paper: will work on this ]

Other researchers have focused on Poisson's Ratio [Zandt and Ammon, 1995] as an indicator of bulk crustal composition as well as crustal thickness [Durrheim and Mooney, 1991] to constrain models of crustal formation. Durrheim and Mooney [1991] compile crustal thickness data and P-wave velocity data for crustal units divided into geochronological ages. They show that Proterozoic crust is thicker than Archean crust and has a higher P-wave velocity. The crustal thickness data attained in this study also show thicker crusts for Proterozoic units (38.9 km) than Archean units (38.0 km) though this difference is not as drastic as the greater than 5km difference reported by Durrheim and Mooney. Though $Vp$ data was not explicity estimated in this study, the higher seismic velocity ratio in the Proterozoic units (1.76) than in Archean units (1.74) aligns with their argument for a thicker basal layer in Proterozoic crustal units.

In contrast, Zandt and Ammon [1995] report high Poisson's Ratio values for Archean Shields and conclude that no systematic difference between Archaean and Proterozoic crust exists. Their average, transformed into $Vp/Vs$, is 1.84, much higher than the wieghted average for the Canadian Shield of 1.74 computed herein. The Authors also report a low $V_P/V_S$ value for Cenozoic-Mesozoic crust, 1.73 compared to the value calculated here of 1.77. Zandt and Ammon interpret these results showing high values for older crust and lower values for younger crust as indicating two end member scenarios. In the uniformitarian model, a delamination type process operates to remove mafic lower crust during continental collisions and is subsequently followed by a stabilizing process involving basaltic underplating. This implies that younger crust has not yet completed the crustal genesis cycle. In their second scenario, Precambrian cratons evolve through unique processes and younger orogens involve recycling and reworking existing material with little crustal growth.

Whereas Zandt and Ammon show increasing $V_P/V_S$ values with crustal age the estimates computed in this study indicate an opposite trend, that is lower seismic velocity ratios with crustal age. This aligns more closely to the results of Durrheim and Mooney [1991] and seems to provide evidence for an evolution of crustal processes throughout geological time. Though Durrheim and Mooney accept the hypothesis that Proterozoic crust is more prone to basaltic underplating, thereby explaining higher velocities and thicker crust, Roberta Rudnick [1995; Rucknick and Gao, 2003] has argued persuasively that a range of possible processes may explain this data.

@@

>>>>

Mean crustal composition is an important parameter characterizing continental crust that should not be confused with actual lithological or mineralogical state
of the crust. Rather, mean crustal composition is a representative quantity that serves as a constraint on crustal structure and evolution in space and time. Early
attempts at estimating mean crustal composition were largely performed through extrapolations based on the surface sampling of upper crustal rocks [Rudnick and Gao, 2003]. After the plate-tectonic revolution, Taylor [1967, 1977], proposed the 'island arc' or 'andesite' model which posited that bulk crustal composition is equivalent to the average convergent-margin andesite. Several problems with this theory subsequently emerged; for example, island arcs are more basaltic than andesitic and their accretions will lead to basaltic crustal growth and therefore cannot account for the intermediate composition of post-Archean crust. The recognition of these shortcomings led to the development of seismic methods to interrogate middle and lower crustal levels and place constraints on their composition [Christensen and Mooney, 1995; Rudnick and Fountain, 1995]. Unlike these latter studies, the present work adopts a bulk crustal seismic analysis along the lines of Smithson et. al. [1981] that combines $V_P/V_S$ from teleseismic measurements with contraints on $V_P$ from previous active source studies as presented in the crust 1.0 model.

@@

In Figure \ref{fig:lith} we plot our area-weighted seismic parameter estimates for the entire Canadian data set along with area-weighted average estimates for various tectonic subdivisions on a plot $V_P$ versus $V_S$. Also plotted are values for a selection of mineral assemblages considered to be representative of common crustal rocks from the compilation of laboratory measurements by Christensen [1996]. Dashed lines delimit isocontours in $V_P/V_S$. The choice of common shield lithologies is motivated by observations along 30km of paleo-depth represented in the Kapuskasing uplift of the Superior province [Percival and West, 1994]. The Kapuskasing uplift comprises Archean granite-greenstone terranes dominated by granitoid and gneissic rocks, and contains supracrustal assemblages generally metamorphosed from greenschist to amphibolite facies (hence the inclusion of hornblendite and amphibolite) and early plutonic structures comprised largely of tonalites. Tonalites and granodiorites are also part of the TTG (tonalite-trondhjemite-granodiorite) suite which dominates Archaean cratons [Rudnick, 1995]. Seismic velocity values for these lithologies are taken from the work of Christensen [1996] and are quoted at average crustal pressures of 600 MPa.

The Slave, Superior and Churchill provinces share similar mean crustal compositions, with the Slave province being slightly more tonalitic in average composition. These three provinces cluster around the average for the Canadian Shield. The Grenville Province on the other hand differs significantly in mean composition. One interpretation is that the crust in the Grenville province has a mean crustal composition corresponding to a more mafic lower crustal composition than other Shield provinces. This interpretation is consistent with a tectonic model for the the central Grenville Province which suggests that the mid-crust has been completely excised locally [Eaton et. al., 2005; Rivers, 2008]. However, many stations in the South-East of the province also have seismic ratios greater than 1.78, similar to the high values common in the adjacent St. Lawrence Platform (where the geometric mean seismic ratio is 1.79). The average $V_P/V_S$ value for the Grenville Province is 1.77 marginally higher than the average value computed by the study of Eaton et. al. [2005].

We now take the

[ Continue with crustal composition discussion ]

Other researchers have focused on Poisson's Ratio [Zandt and Ammon, 1995] as an indicator of bulk crustal composition as well as crustal thickness and P-wave velocity [Durrheim and Mooney, 1991] to constrain models of crustal formation. Zandt and Ammon [1995] report high $V_P/V_S$ values for Archean Shields and conclude that no systematic difference between Archaean and Proterozoic crust exists. Their average, 1.84, much higher than the wieghted average for the Canadian Shield of 1.74 computed herein. The Authors also report a low $V_P/V_S$ value for Cenozoic-Mesozoic crust, 1.73 compared to the value calculated here of 1.77. Zandt and Ammon interpret these results showing high values for older crust and lower values for younger crust as indicating two end member scenarios. In the uniformitarian model, a delamination type process operates to remove mafic lower crust during continental collisions and is subsequently followed by a stabilizing process involving basaltic underplating. This implies that younger crust has not yet completed the crustal genesis cycle. In their second scenario, Precambrian cratons evolve through unique processes and younger orogens involve recycling and reworking existing material with little crustal growth.

Whereas Zandt and Ammon show increasing $V_P/V_S$ values with crustal age the estimates computed in this study indicate an opposite trend, that is lower seismic velocity ratios with crustal age. This aligns more closely to the results of Durrheim and Mooney [1991] who provide results showing a thick higher velocity Proterozoic crust compared to a thinner lower velocity Archean crust. The crustal thickness data attained in this study also show thicker crusts for Proterozoic units (38.9 km) than Archean units (38.0 km) though this difference is not as drastic as the greater than 5km difference reported by Durrheim and Mooney. Though $Vp$ data was not explicity estimated in this study, the higher seismic velocity ratio in the Proterozoic units (1.76) than in Archean units (1.74) aligns with their argument for an evolution of crustal processes throughout geological time. Though Durrheim and Mooney accept the hypothesis that Proterozoic crust is more prone to basaltic underplating, thereby explaining higher velocities and thicker crust, Roberta Rudnick [1995; Rucknick and Gao, 2003] has argued persuasively that a range of possible processes may explain this data.

@@

Smithson, S. B. (1978) Modeling continental crust-structural and chemical constraints, Geophys. Res. Lett. 5(9), 749-775
